---
title: "Final_document of data cleaning"
author: "Futu Chen"
date: "11/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## This RMD holds my steps of data cleaning from the very beginning

**Please carefully read through the comments before hitting "kint". Make sure that you acknowledge that it costs time running them**  
**All the outputs are saved as a RDS. My other codes will use the cleaned datasets**


## step 1: Read in raw data. Preliminary data cleaning. Lots of tears.
```{r}
"%!in%" = Negate("%in%") #define "not in" - will use multiple times

#bring packages
library(ggplot2)
library(RColorBrewer)
library(tidyverse)
library(readxl)
library(tidyr)
library(stringr)
library(data.table)
library(ggnetwork)
library(intergraph)
library(plyr)
library(plotly)

#try to convert .cl2 race file into editable files
#use 2018 as a try
temp <- read_excel("scy2018 NOSPLIT.xlsx")
temp <- temp[-1,] #get rid of the first line
colnames(temp) <- "mess" #name the only col as mess
#only look for individual events
temp <- subset(temp, startsWith(as.character(mess), "D02")) #D02 indicates individual event
# .cl2 file is organized by character space, so split it by the string length
temp2 <- separate(temp, mess, into = c("eventcode", "Name","USSNUM","othermess","agegender","eventtype","eventnum","agegroup","entrytime","prelim_time","final_time","prelim_heat","prelim_lane","finalheat","finallane","place","point","else"), sep = c(11,39,55,63,67,72,76,88,97,106,124,126,128,130,132,135,138,142,153), remove = FALSE)

#'point' and 'else' need to move one col left
temp2$place <- temp2$point
temp2$point <- temp2[,19]
temp2 <- temp2[,2:18]#drop mess and last col
#split name into last name and first name
temp3 <- temp2  %>%
  separate(Name, c("LastName", "FirstName"), ",")
#split age gender - orginal format is ageMM/ageFF
temp3 <- temp3  %>%
  separate(agegender, c("Age", "Gender"), 2)
names(temp3)
temp3 <- temp3[, c(1,2,3,6,7,8,10,11,13,17,18,19)]
str(temp3)#last column is tricky, throw out!
temp3 <- temp3[,-12]
#save as a csv file
write.csv(temp3, "NESCY2018cleaned.csv")


#####try 2019
#2019
temp <- read_excel("SCY2019 NOSPLIT.xlsx")
temp <- temp[-1,]
colnames(temp) <- "mess"
#only look for individual events
temp <- subset(temp, startsWith(as.character(mess), "D02"))
temp2 <- separate(temp, mess, into = c("eventcode", "Name","USSNUM","othermess","agegender","eventtype","eventnum","agegroup","entrytime","prelim_time","final_time","prelim_heat","prelim_lane","finalheat","finallane","place","point","else"), sep = c(11,39,55,63,67,72,76,88,97,106,124,126,128,130,132,135,138,142,153), remove = FALSE)
temp2$place <- temp2$point
temp2$point <- temp2[,19]
temp2 <- temp2[,2:18]#drop mess and last col
temp3 <- temp2  %>%
  separate(Name, c("LastName", "FirstName"), ",")
temp3 <- temp3  %>%
  separate(agegender, c("Age", "Gender"), 2)
names(temp3)
temp3 <- temp3[, c(1,2,3,6,7,8,10,11,13,17,18,19)]
temp3 <- temp3[,-12]
#check to see if there is only MM and FF, make sure that columns are seperated well
table(temp3$Gender)
write.csv(temp3, "NESCY2019cleaned.csv")


##try#2017
temp <- read_excel("SCY2019 NOSPLIT.xlsx")
temp <- temp[-1,]
colnames(temp) <- "mess"
#only look for individual events
temp <- subset(temp, startsWith(as.character(mess), "D02"))
temp2 <- separate(temp, mess, into = c("eventcode", "Name","USSNUM","othermess","agegender","eventtype","eventnum","agegroup","entrytime","prelim_time","final_time","prelim_heat","prelim_lane","finalheat","finallane","place","point","else"), sep = c(11,39,55,63,67,72,76,88,97,106,124,126,128,130,132,135,138,142,153), remove = FALSE)

temp2$place <- temp2$point
temp2$point <- temp2[,19]
temp2 <- temp2[,2:18]#drop mess and last col
temp3 <- temp2  %>%
  separate(Name, c("LastName", "FirstName"), ",")
temp3 <- temp3  %>%
  separate(agegender, c("Age", "Gender"), 2)
names(temp3)
temp3 <- temp3[, c(1,2,3,6,7,8,10,11,13,17,18,19)]
temp3 <- temp3[,-12]
#check to see if there is only MM and FF, make sure that columns are seperated well
table(temp3$Gender)
write.csv(temp3, "NESCY2017cleaned.csv")

```


After inspecting my .csv files that I made for 2017, 2018 and 2019, I feel confident of letting the loop to take over

```{r}
#Loop over the other years
year <- c(2009:2016)
#place holders
tb <- NULL
tb2 <- NULL
for (i in 1:length(year)) {
  temp <- read_excel(paste0("SCY",year[i]," NOSPLIT.xlsx"))
  temp <- temp[-1,]
  colnames(temp) <- "mess"

  temp <- subset(temp, startsWith(as.character(mess), "D02"))
  temp2 <- separate(temp, mess, into = c("eventcode", "Name","USSNUM","othermess","agegender","eventtype","eventnum","agegroup","entrytime","prelim_time","final_time","prelim_heat","prelim_lane","finalheat","finallane","place","point","else"), sep = c(11,39,55,63,67,72,76,88,97,106,124,126,128,130,132,135,138,142,153), remove = FALSE)
  temp2$place <- temp2$point
temp2$point <- temp2[,19]
temp2 <- temp2[,2:18]#drop mess and last col
temp3 <- temp2  %>%
  separate(Name, c("LastName", "FirstName"), ",")
temp3 <- temp3  %>%
  separate(agegender, c("Age", "Gender"), 2)
names(temp3)
temp3 <- temp3[, c(1,2,3,6,7,8,10,11,13,17,18,19)]
temp3 <- temp3[,-12]
#check to see if there is only MM and FF, make sure that columns are seperated well
 tb <- table(temp3$Gender)
 tb2 <- cbind(tb,tb2)
  write.csv(temp3, paste0("NESCY",year[i],"cleaned.csv"))
}

#create a longitudinal file 
#one year per person per row - long format
year <- c(2009:2019)
all <- NULL
for (i in 1:length(year)) {
  temp <- read.csv(paste0("NESCY",year[i],"cleaned.csv"))
  temp$Year <- year[i]
  all <- rbind(all,temp)
}
table(all$Gender)
str(all)
ID <- c(1:43902)
all$ID <- ID

write.csv(all, "NESCYtenyears.csv", row.names = FALSE)

```


So far I have compiled all the data from 2009 to 2019 into a long format.

However, I need to do some extra works to create the variables of interest:  
1. Time variable: the ultimate goal is to have a "speed" indicator. We may want something that has the unit "yard per second". I will transform the minute:second:milionsecond data into seconds.  
2. Stroke indicator: in the raw data (race file), the event is coded as XXXX1, XXXX2, XXXX3, XXXX4 and XXXX5, where the last number indicates the stroke, and whatever before that indicates the distance   


```{r}
all <- read.csv("NESCYtenyears.csv")
#need to work on eventtype and time data 

####################################### Working with time ###############################
#get rid of the "Y" indicating SCY
all$entrytime <- gsub("Y", "",all$entrytime)
all$final_time <- gsub("Y", "", all$final_time)
#last digit of the event type should be stroke
all <-  mutate(all, eventtypec = as.character(eventtype))
all <- all %>% separate(eventtypec, into = c('distance', 'stroke'), sep = -1, convert = TRUE)
#label strokes
all$stroke.f <- factor(all$stroke, levels=c(1:5), labels = c("Freestyle","Backstroke","Breaststroke","Butterfly","IM"))
#need to work with time

#set milionsecond as its own cell

all2 <- all %>% separate(entrytime, into = c('mmss', 'msec'), sep =-2, remove = FALSE)
all2$mmss <- gsub("[.]", "", all2$mmss)
#time will be in seconds
all2$entrytimev2 <- sapply(strsplit(all2$mmss,":"),
  function(x) {
    x <- as.numeric(x)
    x[1]*60+x[2] #calcuate senconds
  }
)
#this function cannot work for data that is already in seconds
#if there is no minutes, seconds will be NA, so I need to keep seconds as is
all2$entrytimev2 <- ifelse(is.na(all2$entrytimev2 == TRUE), as.numeric(all2$mmss), as.numeric(all2$entrytimev2))
#now add milionsecond back
all2$entrytimev2  <- all2$entrytimev2  + 0.01* as.numeric(all2$msec)

#do same for final time
all3 <- all2 %>% separate(final_time, into = c('mmss', 'msec'), sep =-2, remove = FALSE)
all3$mmss <- gsub("[.]", "", all3$mmss)
#time will be in seconds
all3$finaltimev2 <- sapply(strsplit(all3$mmss,":"),
  function(x) {
    x <- as.numeric(x)
    x[1]*60+x[2] #calcuate senconds
  }
)
#SECONDS will be NA, so I need to keep seconds as is
all3$finaltimev2 <- ifelse(is.na(all3$finaltimev2 == TRUE), as.numeric(all3$mmss), as.numeric(all3$finaltimev2))
all3$finaltimev2  <- all3$finaltimev2  + 0.01* as.numeric(all3$msec)
all3 <- as.data.frame(all3)
check <- all3[c("final_time","finaltimev2")]

#there are some NAs, this is due to the DQ and NS (disqualify and no show)

#YEAH! Done with cleaning! drop cols that I do not want
names(all3)
all3 <- all3[c( "ID" , "eventcode","LastName", "FirstName"  , "Age", "Gender",   "agegroup"   , "entrytime"   ,"final_time" , "finallane"  , "place"   ,    "Year" ,  "ID"  ,   "distance" ,   "stroke"  ,    "stroke.f"  ,  "entrytimev2" ,"finaltimev2")]
all3$Gender <- ifelse(all3$Gender == "MM", "Male", "Female")
saveRDS(all3, file="FinalProjCleaned.Rda")
```


## step 1 continu'd: More cleaning. 

Now I need one more variable:
3. Age group, which is unique to Masters Swimming.


```{r}
dat <- readRDS("FinalProjCleaned.Rda")
club <- read.csv("Team score.csv") 
names(dat)
#group age into age group: swim meets are ranked by age groups : https://www.usms.org/-/media/usms/pdfs/volunteer%20central/rule%20book/2019/part%201.pdf?la=en&hash=DAD58443EFF88175A6816090FA628300DECFF720
#Individual Events—18–24, 25–29, 30–34, 35–39, 40–44, 45–49, 50–54,55–59, 60–64, 65–69, 70–74, 75–79, 80–84, 85–89, 90–94, 95–99, 100–104 
agebreaks <- seq(20,95,5)
#floor to 18 yo
agebreaks[1] <- 18 
agelabels <- c("18-24","25–29", "30–34", "35–39", "40–44", "45–49", "50–54","55–59", "60–64", "65–69", "70–74", "75–79", "80–84", "85–89", "90–94")
setDT(dat)[ , agegroupv2 := cut(Age, breaks = agebreaks, right = FALSE, labels = agelabels)]
#check
check <- dplyr::filter(dat, as.numeric(year) == 2019 & (distance == 100) & (stroke == 4) & (agegroupv2 == "25–29") & Gender == "Male")
sum(duplicated(check$place) == TRUE) #no dupilcations! check success

#add one more column that calc the difference btw seed time and final time (in seconds)

dat2 <- dat %>% mutate(timediff = finaltimev2 - entrytimev2)
summary(dat2$timediff)#842 NAs-checked, either DQ or NS at final
#it may be more reasonable to see diff time by yards!
dat2 <- dat2 %>% mutate(timediff.peryd = timediff/distance)

```

## step 2: Construct network data. Play from long to wide, wide to long.

The data we have now is in long format. If I want to get summary stats for individuals, I need to convert them to wide format.
Additionally, I need some summary statistics by different groups. This will help me to get more variables of interet, such as:  
1. participation numbers: participated years and events in the past;  
2. average places;  
3. with whom they competed with;
4. what their speed looks like comparing all 3582 swimmers;  
5. what their stroke type is;  
etc.

```{r}
library(dplyr)
# returns string w/o trailing whitespace, ref: https://stackoverflow.com/questions/2261079/how-to-trim-leading-and-trailing-whitespace
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
dat2$FullName <- as.character(paste(dat2$LastName, dat2$FirstName,sep=","))
dat2$FullName <- trim(dat2$FullName)

swimmername <- dat2$FullName
uniquename <- as.data.frame(unique(swimmername))
sum(duplicated(uniquename))

 
dat2$eventtype <- as.numeric(paste0(dat2$distance, dat2$stroke))
saveRDS(dat2, file="FinalProjCleanedv2.Rda")

#total number of people in each age group
together <- dat2 %>%  
  group_by(Year, eventtype, agegroupv2) %>%
  summarise(num.swimmer = n()) %>% 
  ungroup()

#number of times participated in the event, regardless of the agegroup
participate_num <- dat2 %>%  
  group_by(eventtype,  FullName) %>%
  summarise(num = n()) %>% 
  ungroup()
colnames(participate_num) <- c("eventtype","FullName" , "num_par" )

place_num <- dat2 %>%  
  group_by(eventtype,  FullName, place) %>%
  summarise(num = n()) %>% 
  ungroup()
colnames(place_num) <- c("eventtype","FullName" , "place"  , "num_place" )


together2 <- dat2 %>%  
  group_by(Year, eventtype, agegroupv2, FullName) %>%
  summarise(num = n()) %>% 
  ungroup()

par_place <- merge(participate_num, place_num, by = c("FullName", "eventtype"))



```


Here I tried some network graph approach. However, the file is too large to share online and too busy to insert in my app.


```{r}
library(plotly)
library(igraph)

#edges
#pair swimmers names. If they competed with each other, then they formed a pair
#by event, year, age group and gender group
#event.list <- unique(dat2$eventtype)
year.list <- unique(dat2$Year)
#agroup.list <- unique(dat2$agegroupv2)
#gender.list <- unique(dat2$Gender)

#get a data that list who competed with who

#If ran once or loaded data, do not run the following loop again!!!
#####################  WARNING   ##############################
#####################TIME CONSUMING!!!####################
#####################  WARNING   ##############################
#####################TIME CONSUMING!!!####################
x <- NULL
for (y in 1:length(year.list)) {
  temp1 <- filter(dat2, Year == year.list[y])
  gender.list <- unique(temp1$Gender)
  for (g in 1:length(gender.list)) {
    temp2 <- filter(temp1, Gender == gender.list[g])
    agroup.list <- unique(temp2$agegroupv2)
     for (a in 1:length(agroup.list)) {
        temp3 <- filter(temp2, agegroupv2  == agroup.list[a])
        event.list <- unique(temp3$eventtype)
         for (e in 1:length(event.list)) {
          # if (unique(temp3$eventtype) %in% event.list) {
           temp4 <- filter(temp3, eventtype  == event.list[e])
           
          #temp4 should have unique names that "competed" (theoritically) in the same group. in reality they do not because competition is arranged by seed time
           #NEED to clearify on the website
           player <- temp4$FullName
           pairs <- as.data.frame(expand.grid(player,player))
           colnames(pairs) <- c("swimmer1","swimmer2")
           pairs$year <- year.list[y]
           pairs$gender <- gender.list[g]
           pairs$event <- event.list[e]
           pairs$agegroup <- agroup.list[a]
           x <- rbind(x, pairs)
          # }#if
        }#e
     }#a
  }#g
}#y

##################### END Of WARNING   ##############################
##################### END OF WARNING   ##############################

#delete duplicats
x<-x[!(x$swimmer1 == x$swimmer2),]
saveRDS(x, file="FinalProjSwimmersCompete.Rda")

x <- readRDS("FinalProjSwimmersCompete.Rda") #this will be the network data I want to use

justswimmer <- x[,c(1,2)]
justswimmer <- ddply(justswimmer,.(swimmer1,swimmer2),nrow) 
check <- filter(x, swimmer1 %in% "Keil, Mark A" & swimmer2 == "Garr, Michael T")#should be 65 obs, checked 

```


I will do more cleaning down the road, but let's stop here and save our data.


-----END of cleaning------
